---
title: "MLB Playoffs Batted Ball Data"
author: "Brendan Karadenes"
format: revealjs
---

```{r}
#/ echo: false
#/ output: false
#/ warning: false
library(baseballr)
library(tidyverse)
library(shiny)
batted_balls <- read.csv("batted_balls.csv")
batted_balls_correa <- batted_balls %>%
  mutate(year = year(ymd(game_date))) %>%
  filter(player_name == "Correa, Carlos")
choices_correa <- batted_balls_correa |> distinct(player_name) |> pull(player_name)
filtered_data_correa <- batted_balls_correa %>%
       mutate(events = case_when(events %in% c("single", "double", "triple", "home_run") ~ events, TRUE ~ "out")) 
```

## Data

* The data was scraped from the baseball savant website

* The data contains 10,121 instances of batted balls for 409 players in the MLB playoffs from 2019-2023

* Variables of interest include exit velocity, distance, coordinates of the batted ball, player name, launch angle, and event type

## Questions of Interest

* How have hitter tendencies changed over time for different players?

* What is the ideal combination of exit velocity and launch angle?

## Spray Chart

* Sample spray chart for player 

* You can filter distance, exit velocity, players and year

* Tells you whether player was a pull, opposite field, or all fields hitter that year

```{r}
ggspraychart(
  data = batted_balls_correa,
  x_value = "hc_x",
  y_value = "-hc_y",
  fill_value = "events",
  fill_palette = c("out" = "black",
                   "single" = "red",    
  "double" = "blue",   
  "triple" = "yellow",  
  "home_run" = "purple"),
  fill_legend_title = "Batter Outcomes\nfor 2019-2023\nMLB Playoffs",
  density = FALSE,
  bin_size = 15,
  point_alpha = 0.75,
  point_size = 2,
  frame = NULL
) 
```


## Exit Velocity vs Distance Correlation

* Sample scatterplot for player for balls in air

```{r}
ggplot(filtered_data_correa, aes(x = launch_angle, y = launch_speed, color = events)) +
      geom_point() +
      labs(x = "Launch Angle (degrees)", y = "Exit Velocity (MPH)", title = "Exit Velocity vs Launch Angle for Carlos Correa") +
      theme_minimal()
```

## Conclusions

* The ideal combination of exit velocity and launch angle is around 100MPH and 25 degrees

* Differences in hitter tendencies vary from player to player



